---
title: "Process PCE Data"
author: "Erik Dean"
date: "1/17/2022"
output: html_document
---

```{r setup, include=FALSE}

# don't output warnings, code, messages in markdown
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = F) 

# import packages we need (and maybe some we don't). 
# run install.packages("") on each of these if you haven't before

library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(lubridate)
library(gtsummary)
library(tidyverse)
library(plotly)
library(reshape)
library(viridis)
library(tibble)
library(forecast)
library(quantmod)
library(fpp3)
library(zoo)
library(vars)
library(WaveletComp)
library(wavethresh)
library(waveslim)

# set color theme for plots

scale_colour_brewer_d <- function(...) {
  scale_colour_brewer(palette = "Dark2", ...)
}

scale_fill_brewer_d <- function(...) {
  scale_fill_brewer(palette = "Dark2", ...)
}

options(
  ggplot2.discrete.colour = scale_colour_brewer_d,
  ggplot2.discrete.fill = scale_fill_brewer_d
)

```

## Description

This RMD processes Personal Consumption Expenditure tables from the BEAs website.  It includes a function to pull from the data with various parameters (e.g. start and end dates, level of items to retrieve, &c.)

## Data Source

Here, we're going to be working with the underlying (the detailed) monthly price and quantity data.  These are tables...
* 2.4.3U. Real Personal Consumption Expenditures by Type of Product, Quantity Indexes
* 2.4.4U. Price Indexes for Personal Consumption Expenditures by Type of Product
* 2.4.5U. Personal Consumption Expenditures by Type of Product

There's an R API to grab these data but the data comes back without indication of the levels of the items, so I've been running this from CSVs from the website.  To do this, go to:

https://apps.bea.gov/iTable/iTable.cfm?1921=underlying&isuri=1&reqid=19&step=2#1921=underlying&isuri=1&reqid=19&step=2

Then to Section 2, then for each table: click Modify on the right and Monthly and Select All Years.  Then click Download on the right and download as CSV.  Save these in your working directory (use getwd() to check what that is) as 243U.csv, 244U.csv, and 245U.csv.

Note: Table 2.4.5U can be used to calculate weights of the items, though I haven't worked on this yet. For more information, see https://www.bea.gov/help/faq/1006 

Import the tables from CSV and format.  This will produce 3 sets of tables for expenditures, prices, and quantities: the normal table of items (exp, pri, qua), a separate table of aggregates, and a separate table for non-profit institutions serving households (NPISH).

```{r import and format tables, echo = T}

#clear the stuff

rm(list = ls())

# import the tables

expenditures <- read.csv("243U.csv", header = F, na.strings = "---")
prices <- read.csv("244U.csv", header = F, na.strings = "---")
quantities <- read.csv("245U.csv", header = F, na.strings = "---")

# format expenditures tables

  # remove title rows (1:4) and footnotes (410:414)
  exp <- expenditures[5:413, ]
  
  # move aggregates to separate table
  aggExp <- exp[c(3, 369:409), ]
  exp <- exp[-c(3, 369:409), ]
  
  # move non-profit institutions serving households (NPISHs) to separate table
  npishExp <- exp[341:367, ]
  exp <- exp[-(341:367), ]
  
  # combine month and year data into 2nd row
  exp[2,3:ncol(exp)] <- as.character(my(paste0(exp[2,3:ncol(exp)], " ", exp[1,3:ncol(exp)])))
  exp <- exp[-1,]
  exp[1, 2] <- "Date"

  # add the date rows to the other tables
  aggExp <- rbind(exp[1,], aggExp)
  npishExp <- rbind(exp[1,], npishExp)

# format prices tables (same as above)

  pri <- prices[5:413, ]
  aggPri <- pri[c(3, 369:409), ]
  pri <- pri[-c(3, 369:409), ]
  npishPri <- pri[341:367, ]
  pri <- pri[-(341:367), ]
  pri[2,3:ncol(pri)] <- as.character(my(paste0(pri[2,3:ncol(pri)], " ", pri[1,3:ncol(pri)])))
  pri <- pri[-1,]
  pri[1, 2] <- "Date"
  aggPri <- rbind(pri[1,], aggPri)
  npishPri <- rbind(pri[1,], npishPri)

# format qua tables (same as above)

  qua <- quantities[5:413, ]
  aggQua <- qua[c(3, 369:409), ]
  qua <- qua[-c(3, 369:409), ]
  npishQua <- qua[341:367, ]
  qua <- qua[-(341:367), ]
  qua[2,3:ncol(qua)] <- as.character(my(paste0(qua[2,3:ncol(qua)], " ", qua[1,3:ncol(qua)])))
  qua <- qua[-1,]
  qua[1, 2] <- "Date"
  aggQua <- rbind(qua[1,], aggQua)
  npishQua <- rbind(qua[1,], npishQua)
  
# remove the original tables
  
rm(expenditures)
rm(prices)
rm(quantities)
  
```

## Query the Data

The following function allows querying the main tables (exp, qua, and pri) based on item levels, where goods and services are level 1; durable goods, nondurable goods, and HH cons exp on services are level 2; and so on.  Often, we'll want the lowest (i.e. most granular level), which can be retrieved with the lowestLevel = T parameter.

```{r make ts function}

makeTimeSeries <- function(df, startDate = "1900-01-01", endDate = "2500-01-01", onlyLowestLevel = T, removeNAs = T){

  # make data params actual dates
  startDate = ymd(startDate)
  endDate = ymd(endDate)
  
  #########################################
  # adjust start and end dates
  
  x <- ymd(df[1, 3:ncol(df)]) # get the dates from the tables

  # if specified start/end date is/are out of bounds, then set it/them to the first/last date
  if (startDate < x[1]) { startDate <- x[1] }
  if (endDate > x[length(x)]) { endDate <- x[length(x)] }
  
  cutLateYears <- ncol(df) - sum(x > endDate)
  
  if(cutLateYears <= ncol(df)){ 
    df <- df[, 1:cutLateYears] 
  }
  
  cutEarlyYears <- sum(x < startDate)
  
  if(cutEarlyYears > 1){ 
    df <- df[,-(3:(cutEarlyYears+2))] 
  }
   
  #########################################
  # remove rows with NAs
  # this will ensure that levels that have sub-levels but not full data for them get marked as the lowest level
  # but those lower levels will be dropped. usually, NAs are due to new products, 
  # so what gets cut out here will depend on the time frame, which is why I've run the start/end dates cut prior to this
  
  if (removeNAs){ df <- na.omit(df) }
  
  ############################################
  ### create a new level system for categories
  
  # find number of preceding spaces (the function tells us position of first non-space, so subtract one from that)
  # (and there are 4 spaces for each additional levels, so divide by 4)
  numSpac <- (regexpr("[A-Z0-9]", df[, 2]) - 1) / 4
  
  n <- 2 # start off on the 2nd row, the first is a header row 
  # current level (higher number means lower sublevel), there won't be a 9th level, it's there so the 2nd loop doesn't throw NAs
  cl <- c(0, 0, 0, 0, 0, 0, 0, 0, 0) 
  upLvl <- 0 # goes up by one if we move back to a higher category level
  lp <- 1 # current position in level
  
  level <- vector("list", nrow(df)) # this will be the actual level code
  level[[1]] <- "Level"
  lowestLevel <- vector("list", nrow(df)) # this will indicate if the current level has no sub-levels
  lowestLevel[[1]] <- "lowestLevel"

  while (n <= nrow(df)){
    if (numSpac[n] > numSpac[n-1]){ # move to sublevel, set previous row's lowestLevel to 0, and zero out upLvl
      lp <- lp + 1  
    }
    else if (numSpac[n] < numSpac[n-1]){ # move to higher level
      lp <- lp - (numSpac[n-1] - numSpac[n]) # determine level position based on how far we dropped back
      cl[lp+1:8] <- 0 # zero out levels to right of this higher level

    }
    cl[lp] <- cl[lp] + 1
    
    level[n] <- paste(cl[1], cl[2], cl[3], cl[4], cl[5], cl[6], cl[7], cl[8], cl[9], sep = ".")
    n <- n + 1
  }
  
  # add the level column to the tables
  df <- add_column(df, "Level" = level, .after = 1)
  
  n <- 2
  while (n < nrow(df)){  
    # if the first zero in the current row appears earlier than in the next row, then the next item is a sublevel and this one isn't the lowest
    if (str_locate(df$Level[n], "0")[1] < str_locate(df$Level[n+1], "0")[1]){
      lowestLevel[n] <- 0
    } else { lowestLevel[n] <- 1 }
    n <- n + 1
  }
  # the loop won't pick up the last the row so do that manually
  # it should always be 'other household services', which is a lowest level
  lowestLevel[nrow(df)] <- 1
  
  # add the lowestLevel column to the tables
  df <- add_column(df, "lowestLevel" = lowestLevel, .after = 2)
  
  # move first row to column names
  names(df) <- as.character(unlist(df[1,]))
  colnames(df)[2] <- "Category"
  
  # create bridge between level number and product category
  if (onlyLowestLevel) { df  <- df[df$lowestLevel == 1, ] }
  bridgeLvlCat <- df[,c(1, 2, 4)]
  df <- df[-1,]
  bridgeLvlCat <- bridgeLvlCat[-1,]
  
  # make level numbers row names
  row.names(df) <- df[,2]

  # transpose and pare down the base tables to only include the lowest levels
  # if a category has sub-levels those are included but not the higher level (that the lower ones aggregate to)

  df <- as.data.frame(t(df[,-(1:4)]))
  df[,] <- as.numeric(unlist(df[,]))

  # make the time series
  df.ts <- ts(df, start = c(substr(startDate, 1, 4), substr(startDate, 6, 7)), end = c(substr(endDate, 1, 4), substr(endDate, 6, 7)), freq = 12)

  # remove stuff
  #rm(level, cl, lp, n, numSpac, upLvl, lowestLevel, cutEarlyYears, cutLateYears, startDate, endDate, x, onlyLowestLevel, removeNAs)

  output <- list(item = df, level = as.character(bridgeLvlCat[, 2]), name = trimws(bridgeLvlCat[, 3]), 
                 line = bridgeLvlCat[, 1], startDate = startDate, endDate = endDate)

  return(output)
}

```

Below is some sample code for ploting a selection from the table.

```{r plot a series}

# get the table from the function
priSeries <- makeTimeSeries(pri)

# selection an item by number
i <- 130

# make it a time series (for some reason I could never get it to come out a ts from the function)
series <- ts(quaSeries$item[i], start = as.numeric(substr(quaSeries$startDate, 1, 4)), end = as.numeric(substr(quaSeries$endDate, 1, 4)), freq = 12)

# plot the series
plot(series, main = quaSeries$name[i], ylab = "Price")

```

Export the tables to CSV.

```{r save tables to csv}

# get the table from the function
priSeries <- makeTimeSeries(pri)
quaSeries <- makeTimeSeries(qua)
expSeries <- makeTimeSeries(exp)

p <- priSeries$item
colnames(p) <- priSeries$name 

q <- quaSeries$item
colnames(q) <- quaSeries$name

e <- expSeries$item
colnames(e) <- expSeries$name

write.csv(p, "prices.csv")
write.csv(q, "quantities.csv")
write.csv(e, "expenditures.csv")

rm(p, q, e, priSeries, quaSeries, expSeries)

```


